Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-14 aistats.tex
Write about impact

%\textcolor{red}{This comes a bit from nowhere. Maybe phrase it like (its not sexy wirting here, just the idea of the structure) "The measure of closeness between theta and its estimate hat theta is an active area of research. We distinguish two way to do that: 1) distance in the parameter space and 2) distance between their associated distribution. In the first case, the analysis use norms blabla. In the second, we use KL"}
{\bf Metrics.}
Once we have an estimator $\hat \nat$, we wish to assess its quality:
how close is it from some optimal $\nat^*$.
This is necessary, for instance to choose between 2 estimators (model selection).
We distinguish two ways of doing that.
{\bf 1)} distance in parameter space $d(\nat,\nat^*)$. This is the focus of \emph{parameter estimation}, yielding results such as the asymptotic efficiency of the MLE via the Cramer-Rao lower-bound \citep{aitken1942estimation}.
{\bf 2)} distance between distributions, as studied in \emph{density estimation}.
For this purpose, the Kullback-Leibler (KL) divergence $\KL\paren{p_{\nat^*} || p_{\nat} }$  arises naturally from information theory,
but its lack of robustness to misspecification\footnote{
$\KL(p||q) = +\infty$ if the model $q(x)=0$ for some $x$ from the true distribution$p(x)>0$.
}
has led statisticians to study other distances, such as the $L^2$ norm \citep[e.g. \S1.2]{tsybakov2009introduction}.
As we will see, with exponential families, the KL divergence is also a Bregman divergence between parameters, thus drawing a connection between these two lines of research.
Hence the natural question:

%\textcolor{red}{The sample complexity of the quality of the estimator is well known when we use norms between parameters, in opposition to the case where we measure the distance between distribution. }
% RLP : that is not so true. the citation tsybakov does gie bound for non-parametric estimation with kernels in L^2 distance.

%\begin{center}
%	\framebox{
%		\parbox[c][]{0.9\linewidth}{
%			\begin{center}
%			\emph{Can we upper bound  the expected value of the KL divergence for MLE or MAP?}
%		\end{center}
%		}
%	}
%\end{center}

\begin{equation}
\boxed{
\begin{aligned}
	\textit{Can we upper}&\textit{ bound  the expected value of} \\
	&\KL(p_{\nat^*} || p_{\hat \nat_n^\text{MLE/MAP} })?
\end{aligned}
}
\tag{$\star$}
\label{problem}
\end{equation}
% RLP : that is a dirty hack, but I would like to be able to reference the problem somehow.

We know asymptotic answers.
We also know simple upper bounds if $\logpart$ is quadratic -- e.g. $X$ is Gaussian with known variance --
or close to quadratic (see \S\ref{ssec:quadratic}).
But in general, this problem is difficult. In this paper, we show some partial solutions, and give some idea on how to solve it.
%\textcolor{red}{Maybe say that "in this paper, we show that this problem is difficult and we give the reason why and some idea on how to solve it"}
%RLP poroblem : we don't show it is difficult, neither do we explain why it is difficult. That sounds hard to do.

@@

\textcolor{red}{I would emphasize more that you are talking about the impact of the work rather than contribution. Also I think its better to write this before the contribution, to motivate better why we did this work. }A solution to this problem would allow the field of optimization tackle a wider class of generalized linear models than fixed variance gaussians (linear regression) and Bernoulli (logistic regression).
It could also help develop accelerated optimization methods.

\paragraph{Contributions}
After formalizing the problem of upper bounding the KL of MAP estimators (\S\ref{sec:problem}), along with its asymptotic properties (\S\ref{ssec:asymptote}), we make the following contributions.
\begin{itemize}
	\itemsep0em
	\item We highlight sufficient conditions to characterize when a (local) quadratic approximation of the KL is valid, offering a partial answer to~\eqref{problem} (\S\ref{ssec:quadratic}-\ref{ssec:local-quadratic}).
	\item We investigate a generalized bias-variance decomposition of the KL\citep{pfau2013generalized}. \TODO{revealing what ? the ideal shape of a rate and a sub-problem} (\S\ref{ssec:bias-variance})
	\item We answer~\eqref{problem} with an upper bound in the special case of a Gaussian with known mean but unknown variance $\cN(0,\sigma^2)$ (\S\ref{ssec:gaussian-variance}) .
	\item We establish a connection between~\eqref{problem} and optimization, and use it to highlight areas for progress in the stochastic optimization of barrier objectives (\S\ref{sec:optimization}).
\end{itemize}

\paragraph{Notation}
$X,\,T(X)$ and $T$ \textcolor{red}{Don't we have T = T(X), but written tuch that it's simpler? I think its better to say it that we write T instead of T(X) for simplicity, rather than saying its a random variable.} are random variables, while $x$ is a sample.
$n$ is the number of samples and $d= \dim(T)$.
$\langle \cdot , \cdot \rangle$ is the Euclidean scalar product in $\real^d$.

>>>>

%\textcolor{red}{This comes a bit from nowhere. Maybe phrase it like (its not sexy wirting here, just the idea of the structure) "The measure of closeness between theta and its estimate hat theta is an active area of research. We distinguish two way to do that: 1) distance in the parameter space and 2) distance between their associated distribution. In the first case, the analysis use norms blabla. In the second, we use KL"}
{\bf Metrics.}
Once we have an estimator $\hat \nat$, we wish to assess its quality:
how close is it from some optimal $\nat^*$.
This is necessary, for instance to choose between 2 estimators (model selection).
We distinguish two ways of doing that.
{\bf 1)} distance in parameter space $d(\nat,\nat^*)$. This is the focus of \emph{parameter estimation}, yielding results such as the asymptotic efficiency of the MLE via the Cramer-Rao lower-bound \citep{aitken1942estimation}.
{\bf 2)} distance between distributions, as studied in \emph{density estimation}.
For this purpose, the Kullback-Leibler (KL) divergence $\KL\paren{p_{\nat^*} || p_{\nat} }$  arises naturally from information theory,
but its lack of robustness to misspecification\footnote{
$\KL(p||q) = +\infty$ if the model $q(x)=0$ for some $x$ from the true distribution$p(x)>0$.
}
has led statisticians to study other distances, such as the $L^2$ norm \citep[e.g. \S1.2]{tsybakov2009introduction}.
As we will see, with exponential families, the KL divergence is also a Bregman divergence between parameters, thus drawing a connection between these two lines of research.
Hence the natural question:
%\textcolor{red}{The sample complexity of the quality of the estimator is well known when we use norms between parameters, in opposition to the case where we measure the distance between distribution. }
% RLP : that is not so true. the citation tsybakov does gie bound for non-parametric estimation with kernels in L^2 distance.
%\begin{center}
%	\framebox{
%		\parbox[c][]{0.9\linewidth}{
%			\begin{center}
%			\emph{Can we upper bound  the expected value of the KL divergence for MLE or MAP?}
%		\end{center}
%		}
%	}
%\end{center}

\begin{equation}
\boxed{
\begin{aligned}
	\textit{Can we upper}&\textit{ bound  the expected value of} \\
	&\KL(p_{\nat^*} || p_{\hat \nat_n^\text{MLE/MAP} }) \ ?
\end{aligned}
}
\tag{$\star$}
\label{problem}
\end{equation}
% RLP : that is a dirty hack, but I would like to be able to reference the problem somehow.

We know asymptotic answers.
We also know simple upper bounds if $\logpart$ is quadratic -- e.g. $X$ is Gaussian with known variance --
or close to quadratic (see \S\ref{ssec:quadratic}).
But in general, this problem is difficult. In this paper, we show some partial solutions, and give some ideas on how to solve it.
%\textcolor{red}{Maybe say that "in this paper, we show that this problem is difficult and we give the reason why and some idea on how to solve it"}
%RLP poroblem : we don't show it is difficult, neither do we explain why it is difficult. That sounds hard to do.

@@

{\bf Impact.}
While a solution to the fundamental problem~\eqref{problem} would be of independent interest -- e.g. assessing the importance of the prior in MAP with few samples-- we believe it would have repercussions to other fields. 
For instance, it could enable stochastic optimization to tackle a wider class of generalized linear models than fixed variance gaussians (linear regression) and Bernoulli (logistic regression) \citep{bach2013nonstronglyconvex}.
%Or it could help developing accelerated (stochastic?) algorithms for barrier objectives  overcoming the limitations of non-Euclidean geometry \citep{dragomir2021optimal}. \rlp{Fred I need your feeedback on that one.}

% outcomes
% quantifying the importance of the prior, and few sample behavior, for instance in Gaussians
% stochastic algorithms for barrier objectives
% such as a lot of GLM
% accelerated methods ?

{\bf Contributions.}
After formalizing the problem~\eqref{problem} (\S\ref{sec:problem}), along with its asymptotic properties (\S\ref{ssec:asymptote}), we make the following contributions.
\begin{itemize}
	\itemsep0em
	\item We highlight sufficient conditions to characterize when a (local) quadratic approximation of the KL is valid, offering a partial answer to~\eqref{problem} (\S\ref{ssec:quadratic}-\ref{ssec:local-quadratic}).
	\item We investigate a generalized bias-variance decomposition of the KL\citep{pfau2013generalized}. \TODO{revealing what ? the ideal shape of a rate and a sub-problem} (\S\ref{ssec:bias-variance})
	\item We answer~\eqref{problem} with an upper bound in the special case of a Gaussian with known mean but unknown variance $\cN(0,\sigma^2)$ (\S\ref{ssec:gaussian-variance}) .
	\item We establish a connection between~\eqref{problem} and optimization, and use it to highlight areas for progress in the stochastic optimization of barrier objectives (\S\ref{sec:optimization}).
\end{itemize}

{\bf Notation.}
$X$ and $T=T(X)$ are random variables, while $x$ is a sample.
$n$ is the number of samples and $d= \dim(T)$.
$\langle \cdot , \cdot \rangle$ is the Euclidean scalar product in $\real^d$.
